---
title: "Explatory Data Analysis on Datasets"
author: "Lance Salmingo"
date: "23-05-2025"
output:
  html_document:
    code_folding: show
    toc: true
    toc_float: true
    theme: flatly
    highlight: haddock
---

For this project you would need to install the following packages:
```
install.packages("tidyverse")
install.packages("ggplot2")
install.packages("conflicted")
install.packages("lubridate")
install.packages("ggpubr")
```

# Setup
```{r setup}
# Load libraries

library(tidyverse)  # For data wrangling
library(ggplot2)    # For visualisation
library(conflicted) # For handling library conflicts
library(lubridate)  # For parsing date values
library(ggpubr)     # For stat_cor()
library(readxl)     # For reading excel files

# Load functions to create input files for ORYZA
source("ORYZA_Model_RTOOLS/Make_EXP_ORYZA.R")
source("ORYZA_Model_RTOOLS/Make_WTH_ORYZA.R")

# Prefer dplyr
conflict_prefer("filter",  "dplyr")
conflict_prefer("select",  "dplyr")
conflict_prefer("lag",     "dplyr")
conflict_prefer("summarise", "dplyr")

# Knitr options
knitr::opts_chunk$set(echo = TRUE, error = TRUE, message = FALSE)
```

# Read datasets
```{r message=FALSE}
# Read csv files
iot_data <- read_csv("./datasets/IoTProcessed_Data.csv")
res_data <- read_excel("./ORYZA_v3.5/ORYZA Analysis Tool(v4.2).xlsm", 
                       sheet = "res", skip = 1)
op_data <- read_excel("./ORYZA_v3.5/ORYZA Analysis Tool(v4.2).xlsm", 
                      sheet = "op")

weather_data <- read_csv("./datasets/weather_input.csv")

# Fix column naming
iot_data <- iot_data %>%
  rename(
    temperature = tempreature
  )

iot_data
```

# Explore IoT Dataset

```{r, message=FALSE, warning=FALSE}
# Since ORYZA takes daily values we get group by day and get the mean
# values of each parameter
iot_daily <- iot_data %>%
  # 1) Extract the calendar day from your POSIXct
  mutate(day = as.Date(date)) %>%
  # 2) Group by that day
  group_by(day) %>%
  # 3) Average every numeric column except the original date
  summarise(
    min_temp = min(temperature, na.rm = TRUE),
    max_temp = max(temperature, na.rm = TRUE),
    mean_h = mean(humidity, na.rm = TRUE),
    mean_n = mean(N, na.rm = TRUE),
    mean_p = mean(P, na.rm = TRUE),
    mean_k = mean(K, na.rm = TRUE)
  ) %>%
  ungroup()

# Inspect
print(iot_daily)

# Plot Temperature and Humidity
ggplot(iot_daily, aes(x = day)) +
  # ribbon for temperature
  geom_ribbon(aes(ymin = min_temp,
                  ymax = max_temp),
              fill  = "firebrick",
              alpha = 0.5) +
  # line for humidity, but we’ll rescale it to sit on the same primary axis
  geom_line(aes(y = mean_h * (max(iot_daily$max_temp)/100)),
            color = "steelblue",
            size  = 1) +
  
  # primary axis is temperature
  scale_y_continuous(
    name     = "Temperature (°C)",
    # mirror secondary axis by undoing the scale factor
    sec.axis = sec_axis(~ . * (100 / max(iot_daily$max_temp)),
                        name = "Humidity (%)")
  ) +
  
  labs(
    x     = "Date",
    title = "Daily Min–Max Temperature & Mean Humidity"
  ) +
  theme_minimal() +
  theme(
    axis.text.x     = element_text(angle = 45, hjust = 1),
    legend.position = "none"
  )


# 1) Pivot to long form
daily_long <- iot_daily %>%
  pivot_longer(
    cols      = c(mean_n, mean_p, mean_k),
    names_to  = "Nutrient",
    values_to = "Concentration"
  )

# 2) Overlayed line plot
ggplot(daily_long, aes(x = day, y = Concentration, color = Nutrient)) +
  geom_line(size = 0.8) +
  scale_x_date(
    date_labels = "%b %Y",    # e.g. "Nov 2023"
    date_breaks = "1 month"
  ) +
  labs(
    title = "Daily Mean N, P & K Over Time",
    x     = "Date",
    y     = "Mean Concentration",
    color = "Nutrient"
  ) +
  theme_minimal() +
  theme(
    axis.text.x     = element_text(angle = 45, hjust = 1),
    legend.position = "bottom"
  )

```
The red ribbon is the minimum and max temperature within the day and the blue line is the humidity.

# Using RTools to create input files

Inspecting the sample of weather and soil data
```{r}
head(weather_data)
```
From here we can see the daily values for maximum and minimum temperature within
the day. As well as rain in mm, solar radiation, and relative humidity.

Solar radiation and rain can be gathered from other IoT sensors but as of the
current dataset we have, we can use the dataset to construct the same table
to get the max-min temperature and humidity

```{r}
head(iot_daily)

iot_input <- iot_daily %>%
  mutate(
    day = format(day, "%m/%d/%Y")
  ) %>%
  rename(
    DATE = day,
    TMAX = max_temp,
    TMIN = min_temp,
    RHUM = mean_h
  )

iot_input
```
Here we can see that from the IoT data we can replicate the needed data to
create the input for weather.

For the demonstration we would use the sample data weather_data.
Assuming that these were generated from IoT sensors

```{r warning=FALSE, message=FALSE}
head(weather_data)
make_wth_oryza(weather_data, "./datasets", "LBPH", 5.125, 119.875, 29.96, 1)


wth_input <- read_delim("./datasets/WTH/LBPH1.013",
                 delim     = ",",
                 col_names = FALSE)

wth_input
```
As seen here the sample weather_data has been converted to the input for the
model as wth_input.

This will convert the weather data into a format that ORYZA will take as input.


# Explore res dataset
This dataset contains the daily output of the model during the simulation.
TIME      = Day number
CROPSTA   = growth status (1 for sowing, 
                           2 for seed bed,
                           3 for transplanting,
                           4 for main field)
TMIN      = Daily min temperature
TMAX      = Daily max temperature
SON       = Total soil organic nitrogen
WRR14     = Grain yield 14% of moisture
```{r}
head(res_data)
names(res_data)

res <- res_data %>%
  mutate(across(everything(), as.numeric))

res
res %>%
  pivot_longer(
    cols      = c(TMIN, TMAX, SON, WRR14),
    names_to  = "Metric",
    values_to = "Value"
  ) %>%
  ggplot(aes(x = TIME, y = Value)) +
    geom_line() +
    facet_wrap(~ Metric, scales = "free_y", ncol = 1) +
    labs(
      x     = "Day Number",
      y     = NULL,
      title = "TMIN, TMAX, SON & WRR14 Over Time"
    ) +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
```
# Explore op dataset
This dataset contains values that are summarised in the season.
Some of the values are:
EMD     = Julian day of emergence
DAE     = Days after emergence
WRR14   = Grain yield with 14% moisture
RAINCUM = Accumulative rainfall in main field
IRCUM   = Accumulative irrigation in main field
SOC     = Soil organic carbon content at the end of season
SON     = Soil organic nitrogen content at the end of season
S_CO2C  = Total CO2 loss
S_N2ON  = Total N2O loss



```{r}
names(op_data)

head(op_data)

op_data %>%
  select(
    EMD, DAE, WRR14, RAINCUM, IRCUM, SOC, SON, S_CO2C, S_N2ON
  )
```
The model has much more variables which provide information about the 
quality of yield and performance of the crop during the season.

